---
title: "Practical Fall 2019"
author: "Maira Gajda"
date: "10/28/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/UMich_Bio201_F19/")
```

As with each lab, add "completed" to the Rmd filename; update the name in the author line above; update the location of the working directory as necessary. 

# Load packages
```{r Load packages, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
set.seed(7)
```

# Part 1 - concepts 
Type the answers to questions 1-5 in the space below. 

### Question 1 DONE 

Given vector v1 below, will it be read as character or numeric? Why does this happen? [1 point]
```{r}
v1 <- c("one", "2", "three", "4")
```

Vector v1 will be read as characters becasue each value of the vector has "", which indicate a string of characters in R. 

### Question 2 DONE

dataframe1 has 323 rows and dataframe2 has 345 rows. After an inner join the resulting data fame has 312 rows. Why does the result have fewer rows? [1 point]

Using an inner join in R will return only the rows of dataframe1 that have a matching key in dataframe2. If the inner join of the dataframe 1 and 2 results in fewer rows, there were oly 312 rows that had a matching key. 

### Question 3 DONE

What are the three assumptions for a t-test? Explain what to do if the variance assumption is violated. Explain what to do if the normality assumption is violated. [3 points]

The first assumption is that the plotted data follows a normal distribution. The second assumption is that the sample size (n) for the data is sufficently large (>30). The last assumption is that the both  groups have roughly equal variances, meaning that they are homogeneous. 

If the normality assumption is violated, the data could be transformed using a mathematical functions, such as taking the log or ln of the data. This might change the shape of the data to fit a normal distrubution. But since the t-test is a comparison of means, its very important that the data is normal, so it might be best use another statistical test entirely. 

### Question 4 DONE 

You have a data frame (called butyrate_df) with the columns: "Participant_ID", "Study_week", "Semester", "Supplement_consumed", "Avg_butyrate". If this data frame is piped to each of the functions below, evaluate if the description of the results is true or false. [3 points]
```{r eval=FALSE, include=FALSE}
butyrate_df %>%
  filter(Semester == "Fall2019")
# Result1 = keeps only data from Fall 2019 semester

butyrate_df %>%
  filter(Avg_butyrate <= 47.0)
# Result2 = keeps only values equal to 47 mmol/kg 

butyrate_df %>%
  group_by(Semester) %>%
  summarize(New_column = median(Avg_butyrate))
# Result3 = new_column contains the median of each participant 
```

The first function will evaluate to true because the filter function will only selct values in the "Semester" column that are equivalent to "Fall2019". 

The second fucntion will evaluate to false. The comparison operator <= is used, so the fitler function will only return values that are less than or equal to 47.0. 

The last function will evaluate to false. The group_by function will group the data by semester, and then we will find the median of the avg_butyrate column in each semester, not by each participant. 

### Question 5 DONE 

something about NAs

The data frame (called breath_df) has the columns: "Participant_ID", "Study_week", "Semester", "Supplement_consumed", "Methane", "Hydrogen", "Carbon_dioxide". This data frame is used for the code below, explain what will happen to any NAs in each of the shown calculations, and how the NAs will effect the underlying data. [2 points]
```{r eval=FALSE, include=FALSE}
# Option 1
breath_df %>%
  filter(!is.na(Methane)) %>% 
  group_by(Participant_ID, Study_week) %>%
  summarize(Avg_methane = mean(Methane))

# Option 2
breath_df %>%
  group_by(Participant_ID, Study_week) %>%
  summarize(Avg_methane = mean(Methane, na.rm = TRUE))
```

NA represents missing data. If there was an issue in the data collection, its best to omit those values entirely, as opposed to counting them as 0, which would affect the statistical tests. 

! is the "not" operator. In Option 1, !is.na returns a vector of the values of Methane that are not NA. Afterwards, the remaining data is grouped by Participant_ID and Study_week and then averaged to find the mean. 

na.rm excludes the values of the dataframe that are NA in the following statisitcal tests. In Option 2, the data is first grouped accordingly, and then mean ofthe values of methane are found, exlcuding the values that are equivalent to NA. 

Both options function nearly identically in this scenario. 



# Part 2 - tidy data 

### Question 6

Find the data frame matching your unique name in the GitHub repository. Find the breath gas and taxa abundance data frames. Download and import all into RStudio and use these data frames to complete the rest of the practical. Name the dataframe to match your uniqname, name the breath gas data frame "breath", name the taxa abundance data frame "taxa". [1 point]
```{r}
gajda <- read_delim(file = "Practial/Raw_data/maigai.txt", 
                    delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("","NA") )

taxa <- read_delim(file = "Practial/Raw_data/taxa.txt", 
                    delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("", "NA") )

breath <- read_delim(file = "Practial/Raw_data/breath.txt", 
                    delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("", "NA") )

```



Update column names in all files match the course conventions. [1 point]

```{r}
colnames(gajda) <- tolower(colnames(gajda))

colnames(taxa) <- tolower(colnames(taxa))

colnames(breath) <- tolower(colnames(breath))
```

View the supplement types in your semester. Are they all entered in the same style? If not update them to match. [1 point]
```{r}
select(gajda, supplement_consumed)
```
The supplements of Fall2015 are all in the same style, to my knowledge. 
### Question 7

What are the dimensions of each data frame? Write the results as a comment after the function. [1 point]
```{r}
dim(gajda) # 445x9 

dim(taxa) # 46881x6 

dim(breath) # 2668x5
```

Select a join that will drop as few data points as possible. Join the data frame containing the participant data with the breath gas data. Name this "joined_df". [1 point]
```{r}
?join
joined_df <- right_join(breath, gajda,  all = TRUE)

#there are more participants in the breath data frame, so I want to return all of the values in breath that have a match in gajda
#right join would return more data points than either have independently 
# in this case inner_join < right_join
```

What are the dimensions of the joined data frame? Write the results as a comment after the function. [1 point]
```{r}
dim(joined_df) # 1654x11
```

### Question 8

Calculate the total SCFAs. [1 point]
(I am assuming that this means the total sum of all the SCFAs, for all of the weeks )
```{r}
joined_df %>%  
  select(propionate_mmol_kg, acetate_mmol_kg, butyrate_mmol_kg) %>%
  sum(na.rm = TRUE)
```

Calculate the weekly mean of each SCFA of each participant. Calculate the weekly mean methane and hydrogen levels of each participant. [1 point]
(from the next question, I am assuming that I need to add the averages as new columns)
```{r}
joined_df %>% 
  group_by(participant_id, study_week, supplement_consumed) %>%  #grouping by each participant id, each week,and each individual supplement
  mutate(
        weekly_mean_butytrate = mean(butyrate_mmol_kg, na.rm = TRUE), 
        weekly_mean_butytrate = mean(butyrate_mmol_kg, na.rm = TRUE),
        weekly_mean_acetate = mean(acetate_mmol_kg, na.rm = TRUE), 
        weekly_mean_propinate = mean(propionate_mmol_kg, na.rm = TRUE), 
        weekly_mean_ch4 = mean(ch4, na.rm = TRUE), 
        weekly_mean_h2 =  mean(h2, na.rm = TRUE)
  )   
         
      
    
  
```

What are the dimensions of the data frame with the avearges? Write the results as a comment after the function.
```{r}
dim(joined_df) # 1618 x 16
```

Join this data frame with the one called "taxa". Name the new data frame "all_data". What are the dimensions of this new data frame? [1 point]
```{r}
all_data <- full_join(taxa, joined_df) # no other specs given for joining, full join 
dim(all_data) # 88509x19
```

Save all data frames generated to your GitHub page. [1 point]
```{r}
# export the data to the curated data folder in the Practial and then add/push/commit to github

write_delim(all_data, path = "Practial/curated_dataframes/all_data.txt", delim = "\t", na = "NA", col_names = TRUE)
write_delim(joined_df, path = "Practial/curated_dataframes/joined_data", delim = "\t", na = "NA", col_names = TRUE)
```

# Part 3 - statistics & graphics 

Update all axes labels and figure titles so they are informative and easy to read. Upload final plots to GitHub. When specified, conduct the appropriate tests, and write the statistic (p, F, R value) as a comment. 

### Question 9

If you have more than one type of supplement consumed in your semester, filter for one type to answer Question 9. Plot each SCFA with the study week (weeks 1 and 3 only) on the x-axis. Make each study week a different color. Hint: you may have to use facets, or create individual plots and combine them into a multi-panelled figure. [3 points]
```{r}
vars <- select(all_data, acetate_mmol_kg, butyrate_mmol_kg, propionate_mmol_kg )


```


```{r}
q9_plot_acetate <-  filter(all_data, supplement_consumed == "Arabino") %>%  
  select(study_week, acetate_mmol_kg)

ggplot(data = q9_plot_acetate, 
  aes(x = study_week, y = acetate_mmol_kg, color = study_week) ) + geom_point() + labs(x = "Study Week", y = "acetate mmol per kgs" )
```


```{r}
q9_plot_butyrate <-  filter(all_data, supplement_consumed == "Arabino") %>%  
  select(study_week, butyrate_mmol_kg)

ggplot(data = q9_plot_butyrate, 
  aes(x = study_week, y = butyrate_mmol_kg, color = study_week) ) + geom_point() + labs(x = "Study Week", y = "butyrate mmol per kgs" )
```

```{r}
q9_plot_propionate <- filter(all_data, supplement_consumed == "Arabino") %>%  
  select(study_week, propionate_mmol_kg)

ggplot(data = q9_plot_propionate, 
  aes(x = study_week, y = propionate_mmol_kg, color = study_week) ) + geom_point() + labs(x = "Study Week", y = "propionate mmol per kgs" )
```


q9_plot_acetate <-  filter(all_data, supplement_consumed == "Arabino") %>%  
q9_plot_butyrate <- butyrate_mmol_kg
q9_plot_propionate <- propionate_mmol_kg
  ggplot(
    aes(x = study_week, y =  , color = study_week ) + 
    labs(x = "Study Week", y = "SCFAs" )+
    geom_point() +
    facet_grid(cols = .~,  scales = "free", space = "free", )
  )
# var on the let hand side and vaariable on the rite had side and theyre seperate by a ~ 
#q9_plot + facet_grid(cols = .~vars)
  
```

Based on the data presented in the plot you generated above, identify which statistical test you would need to determine if there was a change in SCFA concentration between weeks 1 and 3 of the study. Check the assumptions appropriate for the test. [3 points]
```{r}
# assumptions

# test 

# result?
```

Based on the results of the test, conclude if there was a difference in SCFAs between weeks. 

<type answer here> 

### Question 10

Is there a correlation between the abundace of hydrogen generators and breath hydrogen? Create a plot that shows the relationship and conduct the appropriate statistical test(s). [2 points]
```{r}
# plot

# test

# result?
```

What can you conclude from these results?

<type answer here> 

Is there a correlation between the abundace of hydrogen consumers and breath methane? Create a plot that shows the relationship and conduct the appropriate statistical test(s). [2 points]
```{r}
# plot

# test

# result?
```

What can you conclude from these results?

<type answer here> 

### Extra credit:

Is the abundace of hydrogen producers and consumers related? Create a plot, check assumptions, conduct the appropriate statistical test(s), and interpret the results. [1 point]
```{r}
# plot

# test(s)
```

<type answer here> 


-----
end